% Deriving I2SB as Annealed Conditional-Tweedie Ascent
% Full-quality source. Compiles with pdflatex / xelatex, or paste into Overleaf.
% The committed PDF (i2sb_derivation.pdf) was rendered without a LaTeX toolchain
% (matplotlib STIX mathtext); recompile this for the polished version.
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}          % \lVert, \coloneqq, \DeclarePairedDelimiter
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage{newtxtext, newtxmath} % Times-like text + math (drop if unavailable)
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue}

\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\eps}{\varepsilon}
\DeclareMathOperator*{\argmax}{arg\,max}

\title{\vspace{-2em}Deriving I2SB as Annealed Conditional-Tweedie Ascent}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-3em}
\begin{center}\itshape A corrected derivation for pure image-to-image translation
(T1 $\to$ T1CE)\end{center}
\vspace{0.5em}
\hrule
\vspace{1em}

We translate a source $y$ (domain A, e.g.\ T1) into a target $x$ (domain B, e.g.\ T1CE),
from paired data $(x,y)\sim p(x,y)$ and with \emph{no} measurement operator. This note
derives the I2SB algorithm as an annealed ascent driven by a conditional-Tweedie residual,
and isolates the one substitution that separates a blurry one-shot regressor from the bridge.

\section{The MAP objective, and why the Bayes split fails}
The translation MAP problem is
\begin{equation}
  x^\star = \argmax_x\; p(x \mid y).
\end{equation}
The usual move is Bayes, $\argmax_x\,[\log p(y\mid x) + \log p(x)]$, which needs an
\emph{analytic} likelihood $p(y\mid x)$. That exists only when $y$ is a measurement of $x$
through a known operator. In pure translation the entire inter-domain relationship is what
we are learning, so a Gaussian stand-in
\begin{equation}
  p(y\mid x) = \N(y; x, \tau)
  \;\;\Longrightarrow\;\;
  \log p(y\mid x) = -\frac{1}{2\tau}\norm{x-y}_2^2
\end{equation}
is not an approximation but a \emph{different problem} (restoration): it pulls $x$ toward
$y$, biasing the synthesized T1CE back toward T1. We therefore drop the factorization and
build a path of densities between the two domains instead.

\section{A homotopy between the two data densities}
Introduce a stochastic interpolant between target and source, indexed by $t\in[0,1]$:
\begin{equation}
  x_t = \mu_0(t)\,x + \mu_1(t)\,y + \sigma(t)\,\eps,
  \qquad \eps \sim \N(0, I),
\end{equation}
with boundary conditions $\mu_0(0)=1,\ \mu_1(0)=0$ (so $x_0=x$) and
$\mu_0(1)=0,\ \mu_1(1)=1$ (so $x_1=y$), and $\sigma$ vanishing at both ends. Its marginals
$p_t$ form a continuous homotopy from $p_0=p(x)$ to $p_1=p(y)$. The Brownian-bridge choice of
$(\mu_0,\mu_1,\sigma)$ recovers I2SB exactly; any other valid interpolant yields a sibling
method (flow / bridge matching).

\section{Conditional Tweedie on the bridge variable (the correction)}
The reverse dynamics carrying $p(y)$ back to $p(x\mid y)$ have a single unknown, the denoiser
\begin{equation}
  D(x_t, t) \coloneqq \E[\,x \mid x_t\,],
\end{equation}
estimated by plain regression, $\min_\theta \E\,\norm{D_\theta(x_t,t) - x}_2^2$. This is the
conditional-Tweedie object, done correctly. Because, given the pairing,
\begin{equation}
  x_t \mid (x,y) \;\sim\; \N\!\big(\mu_0 x + \mu_1 y,\ \sigma^2 I\big),
\end{equation}
$x_t$ is a genuine Gaussian perturbation of $x$, and the residual is a valid score of the
time-$t$ marginal:
\begin{equation}
  D(x_t,t) - x_t \;\propto\; \nabla \log p_t(x_t).
\end{equation}
Contrast the tempting but \emph{wrong} residual $\E[x\mid y] - y$: since $y$ is not a
Gaussian corruption of $x$, it is the score of nothing, and it is \emph{static} --- one
fixed vector, with nothing to anneal. The valid residual is evaluated at the \emph{evolving}
iterate $x_t$, not at $y$.

\section{The annealed update is the ImMAP relaxation step}
Discretizing the reverse dynamics gives the update
\begin{equation}
  x_{t-1} = x_t + h_t\,\underbrace{\big(D(x_t,t) - x_t\big)}_{\text{ascent direction}}
            + \underbrace{\gamma_t\,\eps}_{\text{annealed noise}},
  \qquad x_{t_{\max}} = y,
\end{equation}
a relaxation toward the denoiser output plus annealed noise. It is structurally identical to
the ImMAP solver
\begin{equation}
  x_t = x_t + h_t\,(v_t - x_t) + \gamma_t\,\eps,
\end{equation}
with the two differences that make translation differ from reconstruction:
\begin{enumerate}[label=(\roman*), nosep]
  \item the denoiser $v_t = D(x_t,t)$ is trained on the bridge marginals rather than on
        Gaussian noise, and
  \item the source $y$ is absorbed into $D$ through the interpolant, so no measurement
        operator or data-consistency solve ever appears.
\end{enumerate}

\section{\texorpdfstring{$\E[x\mid y]$}{E[x|y]} is exactly the first step}
Your instinct was almost right: the very first iterate uses
\begin{equation}
  x_{t_{\max}} = y, \qquad D(y, t_{\max}) \approx \E[\,x \mid y\,],
\end{equation}
so $\E[x\mid y] - y$ is precisely the \emph{first} step of I2SB --- the blurry MMSE mean. The
algorithm does not stop there: it re-evaluates $D$ on the refined iterate $x_{t-1}$ and
repeats. That re-evaluation is the anneal, self-consistent only because the bridge trained
$D$ on the whole path, not just at $x_t = y$. In one line:
\begin{quote}
  \emph{I2SB applies the paired regressor to its own progressively refined output, instead of
  once to $y$.}
\end{quote}

\section{MMSE vs.\ MAP (a bookkeeping caveat)}
Tweedie yields the posterior \emph{mean} $\E[x\mid x_t]$, not the mode, so the stochastic
iteration \emph{samples} $p(x\mid y)$ rather than solving $\argmax$. The two coincide only for
symmetric posteriors. For a single deterministic estimate --- the honest stand-in for MAP ---
take the OT-ODE limit $\gamma_t \equiv 0$, a deterministic transport $y \mapsto x$. Keep the
noise for diverse samples; drop it for one canonical T1CE (often the safer clinical choice,
given the risk of hallucinated enhancement).

\begin{center}\rule{0.5\textwidth}{0.4pt}\end{center}

\paragraph{Algorithm (stochastic I2SB sampling).}
\begin{equation*}
\begin{aligned}
  &\textbf{initialize } x = y \\
  &\textbf{for } t = t_{\max}, \ldots, 1:\\
  &\qquad \hat{x} = D(x, t) \hspace{3.2em}\text{\small\# paired regressor at the current iterate}\\
  &\qquad x = x + h_t\,(\hat{x} - x) + \gamma_t\,\eps \\
  &\textbf{return } x \hspace{3.4em}\text{\small\# sample of }p(x\mid y);\ \gamma_t\!=\!0\text{ for the MAP-like transport}
\end{aligned}
\end{equation*}

\end{document}
